---
title: "PCA with NFL Data"
description: |
  Utilizing SVD to do a Principal Component Analysis
author:
  - name: Ryan Plain
    url: {}
date: 08-18-2021
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Distill is a publication format for scientific and technical writing, native to the web.

Learn more about using Distill at <https://rstudio.github.io/distill>.

# Singular Value Decomposition

### Dimensionality Reduction

This was one of the most fascinating concepts to me when I began my graduate program. We processed an image in python and played around with the different resolutions we could provide based on the values provided. 

I wanted to walk through an example from scratch in R. I'll be using open source NFL data provided by **nflreadr** and the open source developers of **nflverse**. You can check out their work at https://github.com/nflverse.


First load in some NFL data from 2014 - 2020 and create some wide receiver stats from the play by play data. In the NFL, wide receivers come in *all shapes and sizes*, with different skill sets. Some work quick underneath routes, some are deep vertical threats. There is a wide variety within one position group, which makes it a great candidate for dimensionality reduction to look at archetypes. 

Data: 

- **rec**: total receptions to filter out players with small sample size
- **air_yards**: distance the ball traveled in the air before reaching them
- **yards_per_target**: total yards / targets 
- **yards_after_catch**: yards after the ball was caught
- **td_rate**: touchdowns per target
- **outside_rec**: binary variable if the receiver caught the ball outside the hashes
- **dist_from_sticks**: where they were targeted relative to distance needed to continue drive

```{r, include=TRUE}
library(tidyverse)
players <- nflreadr::load_player_stats(seasons = c(2014:2020), file_type = 'rds')

rosters <- nflreadr::load_rosters(c(2014:2020))

pbp <- nflreadr::load_pbp(2014:2020, file_type = 'rds')


pbp_roster <- pbp %>% 
  left_join(rosters %>% 
              select(position, season, gsis_id), by = c('fantasy_id' = 'gsis_id', 'season'))
  

wr <- pbp_roster %>% 
  filter(position == 'WR') 
  
wr_stats_full <- wr %>% 
  mutate(outside_pass = ifelse(pass_location != 'middle', 1, 0),
         pass_air_yards = ifelse(is.na(air_yards), 0, air_yards),
         pass_air_yards = ifelse(ydstogo <= 10, pass_air_yards, NA)) %>% 
  group_by(receiver_id) %>% 
  summarise(
    rec = sum(complete_pass),
    air_yards = mean(pass_air_yards, na.rm = T),
    yards_per_target = mean(yards_gained, na.rm = T),
    yards_after_catch = mean(yards_after_catch, na.rm = T),
    td_rate = mean(pass_touchdown),
    outside_rec = mean(outside_pass, na.rm = T),
    dist_from_sticks = mean(pass_air_yards - ydstogo, na.rm = T)
  ) %>% 
  filter(rec > 50) %>% 
  left_join(pbp %>% count(receiver_id, receiver, posteam) %>% group_by(receiver_id) %>% arrange(-n) %>% mutate(rn = row_number()) %>% filter(rn == 1) %>% select(-n, -rn)) %>% 
  relocate(receiver, .before = rec)

wr_stats <- wr_stats_full %>% select(-starts_with('rec'), -posteam) %>% scale()

wr_stats %>% 
  as_tibble()
```

## PCA from scratch

Singular value decomposition is a matrix factorization method that generalizes eigendecomposition of a matrix.

$U$ 

$\Sigma$ is the diagonal matrix containing the eigenvalues

$V$ contains the eigen vectors


```{r, include=TRUE}

U <- svd(wr_stats)$u
D <- svd(wr_stats)$d
V <- svd(wr_stats)$v

DS <- diag(1/svd(wr_stats)$d[1:2])
US <- as.matrix(U[,1:2])
VS <- as.matrix(V[,1:2])


final <- VS %*% DS %*% t(US)

pca_data <- final[1:2, ] %>% MASS::ginv()

# Principal components explained variance
sum(D[[1]], D[[2]])/sum(D)



```

Using $SVD$ to get the principal components, we can use the largest 2 to explain the ~54% of the variance. This allows for a 2 dimensional scatter plot to analyze players. Both Deebo Samuel (SF) and AJ Brown (TEN) stick out from the chart. Samuel gets used almost as a running back in his system, and AJ Brown is a deep vertical threat that also excels with yards after the catch.  

```{r, include=TRUE}

pca_data %>% 
  as_tibble() %>% 
  bind_cols(wr_stats_full %>% select(starts_with("rec"), posteam)) %>% 
  ggplot(aes(V1, V2)) +
  geom_point() +
  ggrepel::geom_text_repel(aes(label = paste(receiver, posteam)), max.overlaps = 10) +
  theme_minimal()

```

Further, I would like to continue this post by using the 2018 player tracking data available through the NFL Big Data Bowl, to create more dimensions than play by play data allows for. For instance, with player tracking data you have all the information if a player was on the field. You can then look at percentage of time they were utilized in the endzone, yards of separation at target, distance traveled in route, etc. 

I want to give thanks to [Luis Serrano](https://www.youtube.com/watch?v=DG7YTlGnCEo) who has an excellent explanation of this. 








